%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%  SECTION 2c — LLM Agentic Patterns      %%%%
%%%%              (Phase 3)                  %%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{LLM Agentic Patterns (Phase~3)}
\label{sec:phase3}

\noindent\colorbox{yellow!40}{\parbox{0.97\linewidth}{%
  \textbf{[PLACEHOLDER — Phase~3 not yet implemented.]}
  Results, figures, and final analysis will replace this section after
  Phase~3 completion. The bullet points below document the planned design.
}}
\medskip

\begin{itemize}
  \item \textbf{Goal:} build a Deep Research Agent that autonomously plans
        sub-topics, retrieves evidence per sub-topic, aggregates findings, and
        writes a structured biomedical report with grounded citations.

  \item \textbf{Architecture (planned) — ReAct-style agentic loop:}
    \begin{itemize}
      \item[--] \textbf{Planner:} topic $\to$ list of sub-topics via LLM
                (\emph{``What are the main dimensions of this biomedical question?''}).
      \item[--] \textbf{Explorer:} for each sub-topic, run the Phase~1+2 pipeline
                (retrieve, rerank, extract supporting sentences).
      \item[--] \textbf{Aggregator:} merge evidence across sub-topics;
                deduplicate citations; verify consistency across sub-claims.
      \item[--] \textbf{Report writer:} compose structured final report with
                section headings, per-claim citations, and a summary.
    \end{itemize}

  \item \textbf{Key design decisions (planned):}
    \begin{itemize}
      \item[--] Agent calls \texttt{retrieval.search(sub\_topic)} and
                \texttt{generation.generate(docs, sub\_topic)} from shared
                \texttt{src/} — no code duplication across phases.
      \item[--] Each ReAct iteration: Reason $\to$ \texttt{SEARCH}
                $\to$ Observe $\to$ \ldots $\to$ Generate.
      \item[--] Stopping criterion: fixed max iterations or LLM confidence
                threshold.
    \end{itemize}

  \item \textbf{Evaluation (planned):}
    \begin{itemize}
      \item[--] Citation accuracy: fraction of agent-cited PMIDs that are
                in the qrels (binary precision).
      \item[--] Sub-topic coverage: do the planned sub-topics collectively
                address the original clinical question?
      \item[--] Comparison against Phase~2 single-shot generation:
                does multi-step planning improve factual coverage?
    \end{itemize}

  \item \textit{[To be completed after Phase~3 implementation.
                All results and analysis will be added here.]}
\end{itemize}
