%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%              INTRODUCTION               %%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
\label{sec:intro}

% ── MOTIVATION & PROBLEM DOMAIN ──────────────────────────────────────────────
\begin{itemize}
  \item LLMs have significantly advanced biomedical NLP tasks: question answering,
        literature summarisation, clinical note processing.
  \item Key challenge: \textbf{hallucinations and confabulations} — LLMs generate
        plausible but unsupported claims, especially dangerous in clinical settings.
  \item Grounding LLM outputs in retrieved evidence is the standard mitigation strategy
        (Retrieval-Augmented Generation, RAG).
  \item TREC BioGen 2024/2025~\cite{biogen2024} formalises this as a shared task: retrieve PubMed
        abstracts, generate a grounded answer, and cite supporting evidence.

  \item \textbf{Task definition:} given a biomedical question (topic + clinical question +
        narrative), retrieve relevant PubMed abstracts and eventually generate a
        paragraph-length answer with sentence-level citations.

  \item \textbf{Why this is hard:}
    \begin{itemize}
      \item[--] Biomedical vocabulary gap: clinical queries use lay or mixed terminology;
                PubMed abstracts use precise scientific notation.
      \item[--] Precision vs.\ recall tension: missing a supporting study is as harmful as
                citing a contradicting one.
      \item[--] Citation-level granularity: the answer must ground every claim, not just
                list references.
      \item[--] Evaluation complexity: requires IR metrics (MAP, NDCG) \emph{and}
                generation quality metrics (citation accuracy, factual consistency).
    \end{itemize}
\end{itemize}

% ── PROJECT STRUCTURE ─────────────────────────────────────────────────────────
\begin{itemize}
  \item \textbf{Phase 1 — Search \& Evaluation (this report):}
    \begin{itemize}
      \item[--] Index 4{,}194 PubMed abstracts in OpenSearch.
      \item[--] Implement and tune 5 retrieval strategies: BM25, LM Jelinek-Mercer,
                LM Dirichlet, dense KNN, and Reciprocal Rank Fusion (RRF).
      \item[--] Evaluate with MAP, MRR, P@10, R@100, NDCG@100 against
                citation-derived relevance judgements.
      \item[--] Best Phase~1 result: \textbf{KNN (MedCPT) NDCG@100 = 0.8077,
                MAP = 0.6143, R@100 = 0.8933} on 33 held-out test queries.
                RRF (BM25+MedCPT, $k$=60) reaches NDCG@100=0.8245.
    \end{itemize}

  \item \textbf{Phase 2 — RAG \& LLM Judges (upcoming):}
    \begin{itemize}
      \item[--] Cross-encoder reranking of retrieved evidence.
      \item[--] LLM-generated paragraph answers with per-sentence citations.
      \item[--] LLM-as-a-judge evaluation of answer quality and citation faithfulness.
    \end{itemize}

  \item \textbf{Phase 3 — Deep Research Agent (upcoming):}
    \begin{itemize}
      \item[--] ReAct-style agentic loop: plan sub-topics, retrieve per sub-topic,
                aggregate evidence, write structured report.
      \item[--] Multi-step reasoning over biomedical literature.
    \end{itemize}
\end{itemize}

% ── CONTRIBUTIONS ─────────────────────────────────────────────────────────────
\begin{itemize}
  \item \textbf{Contributions (Phase 1):}
    \begin{itemize}
      \item[--] Full IR pipeline from raw JSONL corpus to evaluated ranked lists,
                implemented from scratch following TREC and Lab03 protocols.
      \item[--] Systematic hyperparameter tuning with strict train/test discipline
                (5-fold CV on 32 train queries; test set touched once).
      \item[--] Encoder comparison demonstrating that domain-matched MedCPT
                yields $+$0.0501 NDCG@100 over general-domain msmarco-distilbert
                on the test set (0.7576$\to$0.8077, $+$42.8\% relative on train CV).
      \item[--] Post-hoc analyses: statistical significance tests, IDF-based
                difficulty characterisation, Jaccard overlap, confusion matrix
                at P@10, query/document length correlations.
    \end{itemize}
\end{itemize}

% ── PAPER STRUCTURE ───────────────────────────────────────────────────────────
\noindent
This report is organised as follows.
Section~\ref{sec:biogen} describes the BioGen NL Agent pipeline across three phases.
Section~\ref{sec:eval_setup} details the experimental setup (datasets, metrics, protocols).
Section~\ref{sec:eval_results} presents Phase~1 results and discussion.
Section~\ref{sec:conclusions} concludes and outlines future work.
